---
title: "Modeling the Determinants of Diabetes and Prediabetes Using the 2015 CDC Behavioral Risk Factor Surveillance System (BRFSS) Data"
author: "Oluwafunmibi Fasanya"
date: "2025-10-22"
output:
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: true
---

```{r}
# Load required packages
library(dplyr)
library(skimr)
library(kableExtra)
library(tidyr)
library(janitor)
library(gtsummary)
library(ggplot2)
library(broom)
library(patchwork)
library(scales)


```

```{r}
dbts_df <- read.csv("diabetes_012_health_indicators_BRFSS2015 2.csv")
```

```{r}
str(dbts_df)
```

```{r}
summary(dbts_df)
```
```{r}
dbts_df <- dbts_df %>%
  mutate(Diabetes_binary = ifelse(Diabetes_012 == 0, 0, 1))
```

```{r}
# Convert variables to factors with labels

dbts_df <- dbts_df %>%
  mutate(
    Diabetes_binary = factor(Diabetes_binary,
                             levels = c(0, 1),
                             labels = c("No diabetes", "Prediabetes/Diabetes")),

    Diabetes_012 = factor(Diabetes_012,
                             levels = c(0, 1, 2),
                             labels = c("No diabetes", "Prediabetes", "Diabetes")),
    
    HighBP = factor(HighBP, levels = c(0, 1), labels = c("No high BP", "High BP")),
    HighChol = factor(HighChol, levels = c(0, 1), labels = c("No high cholesterol", "High cholesterol")),
    CholCheck = factor(CholCheck, levels = c(0, 1), labels = c("No check in 5 years", "Checked in 5 years")),
    
    Smoker = factor(Smoker, levels = c(0, 1), labels = c("No", "Yes")),
    Stroke = factor(Stroke, levels = c(0, 1), labels = c("No", "Yes")),
    HeartDiseaseorAttack = factor(HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    Fruits = factor(Fruits, levels = c(0, 1), labels = c("No", "Yes")),
    Veggies = factor(Veggies, levels = c(0, 1), labels = c("No", "Yes")),
    HvyAlcoholConsump = factor(HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes")),
    AnyHealthcare = factor(AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes")),
    NoDocbcCost = factor(NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes")),
    DiffWalk = factor(DiffWalk, levels = c(0, 1), labels = c("No", "Yes")),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    
    GenHlth_reversed = 6 - as.numeric(GenHlth),
    
    # Create ordered factor with proper labels
    GenHlth = factor(GenHlth_reversed,
                     levels = 1:5,
                     labels = c("Poor", "Fair", "Good", "Very good", "Excellent"),
                     ordered = TRUE),
    
    Age = factor(Age,
                 levels = 1:13,
                 labels = c("18-24", "25-29", "30-34", "35-39", "40-44",
                            "45-49", "50-54", "55-59", "60-64", "65-69",
                            "70-74", "75-79", "80+"), ordered = TRUE),
    
    Education = factor(Education,
                       levels = 1:6,
                       labels = c("Never attended school or only kindergarten", "Elementary", "Some high school",
                                  "High school graduate", "Some college",
                                  "College graduate"),ordered = TRUE),
    
    Income = factor(Income,
                    levels = 1:8,
                    labels = c("<$10k", "$10k-$15k", "$15k-$20k", "$20k-$25k",
                               "$25k-$35k", "$35k-$50k", "$50k-$75k", "$75k+"),ordered = TRUE)
  )

```

```{r}
summary(dbts_df)
```

```{r}
# Separate categorical and continuous variables
cat_vars <- c("Diabetes_binary", "Diabetes_012", "HighBP", "HighChol", "CholCheck", "Smoker",
              "Stroke", "HeartDiseaseorAttack", "PhysActivity", "Fruits", "Veggies",
              "HvyAlcoholConsump", "AnyHealthcare", "NoDocbcCost","GenHlth", "DiffWalk",
              "Sex", "Age", "Education", "Income")

cont_vars <- c("BMI", "MentHlth", "PhysHlth")

# ---- Categorical Variables Summary ----
cat_summary <- dbts_df %>%
  dplyr::select(all_of(cat_vars)) %>%
  skimr::skim() %>%
  dplyr::select(skim_variable, n_missing, complete_rate, 
                factor.n_unique, factor.top_counts) %>%
  dplyr::rename(
    Variable = skim_variable,
    `Missing (n)` = n_missing,
    `Complete Rate` = complete_rate,
    `Unique Levels` = factor.n_unique,
    `Top Levels (Count)` = factor.top_counts
  )
```

```{r}

# ---- Continuous Variables Summary ----
cont_summary <- dbts_df %>%
  dplyr::select(all_of(cont_vars)) %>%
  skimr::skim() %>%
  dplyr::select(skim_variable, n_missing, numeric.mean, numeric.sd,
                numeric.p0, numeric.p25, numeric.p50, numeric.p75, numeric.p100) %>%
  dplyr::rename(
    Variable = skim_variable,
    `Missing (n)` = n_missing,
    Mean = numeric.mean,
    SD = numeric.sd,
    Min = numeric.p0,
    Q1 = numeric.p25,
    Median = numeric.p50,
    Q3 = numeric.p75,
    Max = numeric.p100
  )

```



```{r}
# ---- Formatted tables ----
cat_summary %>%
  kable("html", caption = "Summary of Categorical Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "center")
```

```{r}
cont_summary %>%
  kable("html", caption = "Summary of Continuous Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "center")

```

```{r}
# ANOVA function for continuous variables
anova_test <- function(data, variable, by, ...) {
  formula <- as.formula(paste(variable, "~", by))
  aov_model <- aov(formula, data = data)
  broom::tidy(aov_model) %>%
    dplyr::slice(1) %>%     # Keep only the main term
    dplyr::select(p.value)
}

# Define continuous variables
cont_vars <- c("BMI", "MentHlth", "PhysHlth")

# APA-style summary table
apa_table <- dbts_df %>%
  tbl_summary(
    by = Diabetes_012,
    type = list(
      all_of(cont_vars) ~ "continuous"
    ),
    statistic = list(
      all_categorical() ~ "{n} ({p}%)",
      all_continuous() ~ "{mean} (±{sd})"
    ),
    missing = "no"
  ) %>%
  add_overall() %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",
      all_of(cont_vars) ~ anova_test
    )
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(
    c("stat_1", "stat_2", "stat_3") ~ "**Diabetes Status**"
  ) %>%
  bold_labels()%>%
  modify_footnote(
    all_stat_cols() ~ "1 n (%); Mean (±SD)",
    p.value ~ "2 Pearson’s Chi-squared test for categorical variables; One-way ANOVA for continuous variables."
  )

apa_table


```
```{r}
apa_flex <- as_flex_table(apa_table)
apa_flex <- flextable::set_table_properties(apa_flex, width = 1, layout = "autofit")

```


```{r}
library(gtsummary)
library(openxlsx)

# Convert gtsummary table to a data frame
apa_df <- as_tibble(apa_table, col_labels = TRUE)

# Save to Excel
write.xlsx(apa_df, "Diabetes_Summary_Table.xlsx", rowNames = FALSE)

```

## Visualizations

```{r}
## Sex: Side-by-Side Bar Plot ---
p_sex <- dbts_df %>%
  count(Sex, Diabetes_012) %>%
  group_by(Sex) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Sex, y = prop, fill = Diabetes_012)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Sex Distribution by Diabetes Status",
       x = NULL, y = "Percentage") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

## Education: Ordered Horizontal Stacked Bar Plot ---
p_edu <- dbts_df %>%
  count(Education, Diabetes_012) %>%
  group_by(Education) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Education, y = prop, fill = Diabetes_012)) +
  geom_col(position = "fill") +
  coord_flip() +   # 🔹 horizontal layout
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Education Level by Diabetes Status",
       x = NULL, y = "Proportion") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = "none")

## Income: Ordered Horizontal Stacked Bar Plot ---
p_income <- dbts_df %>%
  count(Income, Diabetes_012) %>%
  group_by(Income) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = Income, y = prop, fill = Diabetes_012)) +
  geom_col(position = "fill") +
  coord_flip() +   # 🔹 horizontal layout
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = fill_colors) +
  labs(title = "Household Income by Diabetes Status",
       x = NULL, y = "Proportion") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 10),
        legend.position = "bottom")
p_sex
p_edu
p_income
ggsave("Sex_Dist.png", plot = p_sex, width = 8, height = 6, dpi = 300)
ggsave("Edu_Dist.png", plot = p_edu, width = 8, height = 6, dpi = 300)
ggsave("Income_Dist.png", plot = p_income, width = 8, height = 6, dpi = 300)
```

```{r}
p1 <- dbts_df %>%
  count(Age, Diabetes_012) %>%
  ggplot(aes(x = Age, y = n, fill = Diabetes_012)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c( "#27ae60", "#f39c12", "#c0392b" )) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Age Distribution", y = "Percentage", x = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

p1

ggsave("Age_Distribution.png", plot = p1, width = 8, height = 6, dpi = 300)

```

```{r}
sex_age_diabetes <- dbts_df %>%
  group_by(Sex, Age) %>%
  summarise(diabetes_rate = mean(Diabetes_012 != "No diabetes"),
            prediabetes_rate = mean(Diabetes_012 == "Prediabetes"),
            diabetes_only = mean(Diabetes_012 == "Diabetes"),
            .groups = "drop")

p4 <- ggplot(sex_age_diabetes, aes(x = Age, y = diabetes_rate, 
                                    color = Sex, group = Sex)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#e91e63", "#2196f3")) +
  labs(title = "Sex × Age: Diabetes/Prediabetes Rate by Sex Over Age",
       x = "Age Group", y = "Rate of Diabetes/Prediabetes") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
p4

ggsave("SexAge.png", plot = p4, width = 8, height = 6, dpi = 300)
```

```{r}
ses_data <- dbts_df %>%
  group_by(Income, Education) %>%
  summarise(
    diabetes_rate = mean(Diabetes_012 != "No diabetes"),
    n = n(),
    .groups = "drop"
  ) 

p6 <- ggplot(ses_data, aes(x = Income, y = Education, fill = diabetes_rate)) +
  geom_tile(color = "white", size = 1.5) +
  geom_text(aes(label = sprintf("%.1f%%", diabetes_rate * 100)),
            color = "white", fontface = "bold", size = 3) +
  scale_fill_gradient2(low = "#27ae60", mid = "#f39c12", high = "#c0392b",
                      midpoint = 0.12, labels = scales::percent,
                      name = "Diabetes\nRate") +
  labs(title = "The Social Determinants: Income and Education Impact on Diabetes/Prediabetes",
       subtitle = "Lower income + lower education = higher diabetes rates",
       x = "Household Income", 
       y = "Education Level") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))

p6

ggsave("SocialDeterminant.png", plot = p6, width = 8, height = 6, dpi = 300)
```

```{r}
#How Conditions Accumulate
health_cas <- dbts_df %>%
  mutate(risk_count = (HighBP == "High BP") + 
                      (HighChol == "High cholesterol") + 
                      (HeartDiseaseorAttack == "Yes") +
                      (Smoker == "Yes") +
                      (BMI > 30)) %>%
  group_by(risk_count, Diabetes_012) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(risk_count) %>%
  mutate(prop = n / sum(n))

p5 <- ggplot(health_cas, aes(x = factor(risk_count), y = prop, 
                                  fill = Diabetes_012)) +
  geom_col(width = 0.7, color = "white", size = 0.5) +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
            position = position_stack(vjust = 0.5), 
            color = "white", fontface = "bold", size = 4) +
  scale_fill_manual(values = c("#27ae60", "#f39c12", "#c0392b"),
                    name = "Status") +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  labs(title = "How Multiple Risk Factors Compound",
       subtitle = "Risk factors: High BP, High Cholesterol, Heart Disease, Smoking, Obesity (BMI>30)",
       x = "Number of Risk Factors Present", 
       y = "Percentage of Group") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        panel.grid.major.x = element_blank())
p5

ggsave("Health.png", plot = p5, width = 8, height = 6, dpi = 300)
```

```{r}
prevention <- dbts_df %>%
  mutate(
    risk_level = case_when(
      HighBP == "High BP" & HighChol == "High cholesterol" & BMI > 30 ~ "Very High Risk",
      (HighBP == "High BP" | HighChol == "High cholesterol") & BMI > 30 ~ "High Risk",
      HighBP == "High BP" | HighChol == "High cholesterol" | BMI > 30 ~ "Moderate Risk",
      TRUE ~ "Low Risk"
    ),
    # Convert to factor with proper ordering
    risk_level = factor(risk_level, 
                       levels = c("Low Risk", "Moderate Risk", "High Risk", "Very High Risk"))
  ) %>%
  group_by(risk_level, PhysActivity, Diabetes_012) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(risk_level, PhysActivity) %>%
  mutate(prop = n / sum(n)) %>%
  filter(Diabetes_012 == "Diabetes")

p7 <- ggplot(prevention, aes(x = risk_level, y = prop, fill = PhysActivity)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            position = position_dodge(width = 0.7), vjust = -0.5,
            fontface = "bold", size = 4) +
  scale_fill_manual(values = c("#e74c3c", "#27ae60"),
                    name = "Physical Activity\nin Past Month") +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Does Exercise Help? Physical Activity Impact Across Risk Levels",
       subtitle = "Even in high-risk groups, physical activity reduces diabetes prevalence",
       x = "Metabolic Risk Level", 
       y = "Diabetes Rate") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom",
        axis.text.x = element_text(size = 10),
        panel.grid.major.x = element_blank())

p7

ggsave("Exercisehelps.png", plot = p7, width = 8, height = 6, dpi = 300)
```

```{r}
# Select key risk factors
risk_factors <- c("HighBP", "HighChol", "Smoker", "PhysActivity", 
                  "Fruits", "Veggies", "HeartDiseaseorAttack")

# Prepare data
plot_data <- dbts_df %>%
  dplyr::select(Diabetes_012, all_of(risk_factors)) %>%
  pivot_longer(-Diabetes_012, names_to = "Risk_Factor", values_to = "Status") %>%
  filter(Status == "Yes" | Status == "High BP" | Status == "High cholesterol") %>%
  count(Diabetes_012, Risk_Factor) %>%
  group_by(Risk_Factor) %>%
  mutate(
    percentage = n / sum(n) * 100,
    Risk_Factor = case_when(
      Risk_Factor == "HighBP" ~ "High Blood Pressure",
      Risk_Factor == "HighChol" ~ "High Cholesterol",
      Risk_Factor == "HeartDiseaseorAttack" ~ "Heart Disease",
      Risk_Factor == "PhysActivity" ~ "Physical Activity",
      TRUE ~ Risk_Factor
    )
  )

# Create plot
p2 <- ggplot(plot_data, aes(x = reorder(Risk_Factor, percentage), y = percentage, 
                      fill = Diabetes_012)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.7), 
            vjust = 0.5, size = 2.5) +
  scale_fill_manual(values = c( "#27ae60", "#f39c12", "#c0392b"),
                    name = "Diabetes Status") +
  labs(
    title = "Prevalence of Risk Factors by Diabetes Status",
    subtitle = "Comparison of key health indicators between groups",
    x = "Risk Factor",
    y = "Percentage (%)",
    caption = "Data: Diabetes Health Indicators Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  coord_flip()

p2

ggsave("Prevalence_Riskfactor.png", plot = p2, width = 8, height = 6, dpi = 300)
```

```{r}
# Create summary data
heatmap_data <- dbts_df %>%
  count(Age, GenHlth, Diabetes_012) %>%
  group_by(Age, Diabetes_012) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

# Create heatmap
p3 <- ggplot(heatmap_data, aes(x = Age, y = GenHlth, fill = percentage)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", percentage)), 
            color = "white", size = 2.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#27ae60", 
    mid = "#f39c12", 
    high = "#c0392b",
    midpoint = 25,
    name = "Percentage (%)"
  ) +
  facet_wrap(~ Diabetes_012, ncol = 1) +
  labs(
    title = "General Health Status Distribution by Age and Diabetes",
    subtitle = "Darker colors indicate higher percentages within each age group",
    x = "Age Group",
    y = "General Health Status",
    caption = "Data: Diabetes Health Indicators Dataset"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "gray90", color = NA),
    legend.position = "right",
    panel.grid = element_blank()
  )
p3
ggsave("Heatmap.png", plot = p3, width = 8, height = 6, dpi = 300)
```
## Logistic Regression

### Purposeful Selection of Explanatory Variables 

#### Construct an initial main effects model with known important Xs and statistically significant ones

```{r}
#Return these variables to factor
dbts_df <- dbts_df %>%
  mutate(
    GenHlth = factor(GenHlth, ordered = FALSE),
    Age = factor(Age, ordered = FALSE),
    Education = factor(Education, ordered = FALSE),
    Income = factor(Income, ordered = FALSE)
  )

```

```{r}
fit1 <- glm(Diabetes_binary~HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies + 
    HvyAlcoholConsump + AnyHealthcare + NoDocbcCost + GenHlth + 
    MentHlth + PhysHlth + DiffWalk + Sex + Age + Education + 
    Income, data = dbts_df, family = binomial)
#summary(fit1)
```

```{r}
library(car)
Anova(fit1) # likelihood-ratio tests for individual explanatory variables
```



```{r}
#Treating them as numeric
dbts_df <- dbts_df %>%
  mutate(
    GenHlth_num = as.numeric(GenHlth),
    Age_num = as.numeric(Age),
    Education_num = as.numeric(Education),
    Income_num = as.numeric(Income)
  )

```


```{r}
fit2 <- glm(Diabetes_binary ~ HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies + 
    HvyAlcoholConsump + AnyHealthcare + NoDocbcCost + GenHlth_num + 
    MentHlth + PhysHlth + DiffWalk + Sex + Age_num + Education_num + Income_num, 
    data = dbts_df, family = binomial)
#summary(fit2)
```

```{r}
library(car)
Anova(fit2) # likelihood-ratio tests for individual explanatory variables
```


```{r}
anova(fit1, fit2, test="Chisq")
```
Model 1 fits better (also has the least AIC value), thus we use the the ordered variables as a factor instead.


#### Conduct backward elimination and retain a variable if it is significant at a more stringent level or it shows evidence of being a relevant confounder

```{r}
fit_backward <- step(fit1, direction = "backward")
```

```{r}
# Display summary of the final model
summary(fit_backward)
```

#### Check for plausible interactions among variables in the model
```{r}
fit1_int <- update(fit1, . ~ . + BMI:Age + GenHlth:PhysHlth + HighChol:HeartDiseaseorAttack + Sex:Age + BMI:PhysActivity)
```


```{r}
fit_backward <- step(fit1_int, direction = "backward")
```

```{r}
#Final Model 
fit1_int_final <- glm(Diabetes_binary ~ HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies +  HvyAlcoholConsump + NoDocbcCost + GenHlth + PhysHlth + DiffWalk + Sex + Age + Education + Income + BMI:Age + GenHlth:PhysHlth + HighChol:HeartDiseaseorAttack + Sex:Age + BMI:PhysActivity  + HvyAlcoholConsump:Age + Smoker:Age, 
    data = dbts_df, family = binomial)
```

```{r}
summary(fit1_int_final)
```

#### Follow up diagnostic investigation
```{r}
anova(fit1, fit1_int_final, test="Chisq")
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract coefficients, standard errors, and p-values
coef_summary <- summary(fit1_int_final)$coefficients
coef_df <- data.frame(
  variable = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  std_error = coef_summary[, "Std. Error"],
  z_value = coef_summary[, "z value"],
  p_value = coef_summary[, "Pr(>|z|)"]
) %>%
  filter(variable != "(Intercept)") %>%
  mutate(
    # Calculate confidence intervals
    ci_lower = estimate - 1.96 * std_error,
    ci_upper = estimate + 1.96 * std_error,
    # Significance levels
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ ".",
      TRUE ~ "NS"
    ),
    sig_factor = factor(significance, 
                       levels = c("***", "**", "*", ".", "NS"))
  )

# Separate main effects from interactions
coef_df <- coef_df %>%
  mutate(
    effect_type = ifelse(grepl(":", variable), "Interaction", "Main Effect"),
    # Clean variable names for better display
    var_clean = gsub("Yes$", "", variable),
    var_clean = gsub("High ", "", var_clean),
    var_clean = gsub("cholesterol", "Chol", var_clean)
  )

## PLOT 1: Coefficient Plot with Confidence Intervals (Main Effects)
main_effects <- coef_df %>% filter(effect_type == "Main Effect")

p1 <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value), 
                                y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  labs(
    title = "Main Effects on Diabetes Risk (Ordered by Significance)",
    subtitle = "Log-odds coefficients with 95% CI",
    x = "Predictor",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 9)
  )

print(p1)

## PLOT 2: Interaction Effects
interactions <- coef_df %>% filter(effect_type == "Interaction")

p2 <- ggplot(interactions, aes(x = reorder(var_clean, -p_value), 
                                y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  labs(
    title = "Interaction Effects on Diabetes Risk (Ordered by Significance)",
    subtitle = "Log-odds coefficients with 95% CI",
    x = "Interaction Term",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8)
  )

print(p2)

## PLOT 3: Odds Ratios (easier interpretation)
main_effects <- main_effects %>%
  mutate(
    OR = exp(estimate),
    OR_lower = exp(ci_lower),
    OR_upper = exp(ci_upper)
  )

p3 <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value), 
                                y = OR, color = sig_factor)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = OR_lower, ymax = OR_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Odds Ratios for Main Effects (Ordered by Significance)",
    subtitle = "OR > 1: increased diabetes risk | OR < 1: decreased diabetes risk",
    x = "Predictor",
    y = "Odds Ratio (log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 9)
  )

print(p3)

## PLOT 4: Top 20 Most Significant Effects (combined)
top20 <- coef_df %>%
  arrange(p_value) %>%
  head(20) %>%
  mutate(
    OR = exp(estimate),
    direction = ifelse(estimate > 0, "Increases Risk", "Decreases Risk")
  )

p4 <- ggplot(top20, aes(x = reorder(var_clean, -p_value), 
                        y = estimate, fill = direction)) +
  geom_col() +
  geom_text(aes(label = significance), 
            hjust = ifelse(top20$estimate > 0, -0.3, 1.3),
            size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("Increases Risk" = "#D55E00", 
                                "Decreases Risk" = "#0072B2")) +
  labs(
    title = "Top 20 Most Significant Predictors",
    subtitle = "Ranked by statistical significance",
    x = "Predictor",
    y = "Coefficient (log-odds)",
    fill = "Effect Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

print(p4)

## Save all plots to files with appropriate sizes

# Plot 1: Main Effects - tall plot to accommodate many variables
ggsave("plot1_main_effects.png", plot = p1, 
       width = 10, height = 14, dpi = 300, bg = "white")

# Plot 2: Interaction Effects - medium height
ggsave("plot2_interactions.png", plot = p2, 
       width = 10, height = 8, dpi = 300, bg = "white")

# Plot 3: Odds Ratios - tall plot for clarity
ggsave("plot3_odds_ratios.png", plot = p3, 
       width = 10, height = 14, dpi = 300, bg = "white")

# Plot 4: Top 20 Most Significant
ggsave("plot4_top20_significant.png", plot = p4, 
       width = 10, height = 8, dpi = 300, bg = "white")

# Optional: Save as PDF for publication quality
ggsave("plot1_main_effects.pdf", plot = p1, 
       width = 10, height = 14)
ggsave("plot2_interactions.pdf", plot = p2, 
       width = 10, height = 8)
ggsave("plot3_odds_ratios.pdf", plot = p3, 
       width = 10, height = 14)
ggsave("plot4_top20_significant.pdf", plot = p4, 
       width = 10, height = 8)
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract coefficients, standard errors, and p-values
coef_summary <- summary(fit1_int_final)$coefficients
coef_df <- data.frame(
  variable = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  std_error = coef_summary[, "Std. Error"],
  z_value = coef_summary[, "z value"],
  p_value = coef_summary[, "Pr(>|z|)"]
) %>%
  filter(variable != "(Intercept)") %>%
  mutate(
    # Calculate confidence intervals
    ci_lower = estimate - 1.96 * std_error,
    ci_upper = estimate + 1.96 * std_error,
    # Significance levels
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ ".",
      TRUE ~ "NS"
    ),
    sig_factor = factor(significance, 
                        levels = c("***", "**", "*", ".", "NS")),
    effect_type = ifelse(grepl(":", variable), "Interaction", "Main Effect"),
    var_clean = gsub("Yes$", "", variable),
    var_clean = gsub("High ", "", var_clean),
    var_clean = gsub("cholesterol", "Chol", var_clean)
  )

# 🔹 Only keep significant results (p < 0.05)
coef_sig <- coef_df %>% filter(p_value < 0.05)

# Separate main effects and interactions
main_effects <- coef_sig %>% filter(effect_type == "Main Effect")
interactions <- coef_sig %>% filter(effect_type == "Interaction")

# 🔹 Plot 1: Main Effects (Significant Only)
p1 <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value),
                               y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", "*" = "#56B4E9"),
    name = "Significance"
  ) +
  labs(
    title = "Main Effects on Diabetes Risk (Significant Predictors)",
    subtitle = "Log-odds coefficients with 95% CI (p < 0.05)",
    x = "Predictor",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 9)
  )

print(p1)

# 🔹 Plot 2: Interaction Effects (Significant Only)
p2 <- ggplot(interactions, aes(x = reorder(var_clean, -p_value),
                               y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", "*" = "#56B4E9"),
    name = "Significance"
  ) +
  labs(
    title = "Interaction Effects on Diabetes Risk (Significant Predictors)",
    subtitle = "Log-odds coefficients with 95% CI (p < 0.05)",
    x = "Interaction Term",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8)
  )

print(p2)

# Plot 1: Main Effects - tall plot to accommodate many variables
ggsave("p1.png", plot = p1, 
       width = 10, height = 8, dpi = 300, bg = "white")

# Plot 2: Interaction Effects - medium height
ggsave("p2.png", plot = p2, 
       width = 10, height = 8, dpi = 300, bg = "white")
```


```{r}
## PLOT 5: Odds Ratios for Interaction Effects (easier interpretation)
interactions <- interactions %>%
  mutate(
    OR = exp(estimate),
    OR_lower = exp(ci_lower),
    OR_upper = exp(ci_upper)
  )

p5 <- ggplot(interactions, aes(x = reorder(var_clean, -p_value), 
                                y = OR, color = sig_factor)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = OR_lower, ymax = OR_upper), width = 0.3) +
  coord_flip() +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Odds Ratios for Interaction Effects (Ordered by Significance)",
    x = "Interaction Term",
    y = "Odds Ratio (log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8)
  )

print(p5)

# Save the plot
ggsave("plot5_interaction_odds_ratios.png", plot = p5, 
       width = 10, height = 8, dpi = 300, bg = "white")

ggsave("plot5_interaction_odds_ratios.pdf", plot = p5, 
       width = 10, height = 8)
```




```{r}
#Return these variables to factor
dbts_df <- dbts_df %>%
  mutate(
    GenHlth = factor(GenHlth, ordered = FALSE),
    Age = factor(Age, ordered = FALSE),
    Education = factor(Education, ordered = FALSE),
    Income = factor(Income, ordered = FALSE)
  )
```


## Baseline Categorical Logit Model

```{r}
library(VGAM)
fit3 <- vglm(Diabetes_012 ~ HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies +  HvyAlcoholConsump + NoDocbcCost + GenHlth + PhysHlth + DiffWalk + Sex + Age + Education + Income, data = dbts_df, family = multinomial(refLevel = "No diabetes"))
#summary(fit3)
```

```{r}
AIC(fit3)
```

```{r}
fit3_int <- vglm(Diabetes_012 ~ HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies +  HvyAlcoholConsump + NoDocbcCost + GenHlth + PhysHlth + DiffWalk + Sex + Age + Education + Income + BMI:Age + GenHlth:PhysHlth + HighChol:HeartDiseaseorAttack + Sex:Age + BMI:PhysActivity  + HvyAlcoholConsump:Age + Smoker:Age, data = dbts_df, family = multinomial(refLevel = "No diabetes"))


AIC(fit3_int)
```


```{r}
levels(dbts_df$Diabetes_012)
```

```{r}
summary(fit3_int)
```

```{r}
VGAM::lrtest(fit3, fit3_int)
```

```{r}
summary(fit3_int)
```

```{r}
coef(fit3_int, matrix = TRUE)
```



```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(VGAM)

# Extract coefficients from VGAM multinomial model
coef_summary <- summary(fit3_int)@coef3

# VGAM returns coefficients for each response level (Prediabetes and Diabetes)
# We'll process both levels

# Create a data frame from coefficients
coef_df <- data.frame(
  variable = rownames(coef_summary),
  estimate = coef_summary[, "Estimate"],
  std_error = coef_summary[, "Std. Error"],
  z_value = coef_summary[, "z value"],
  p_value = coef_summary[, "Pr(>|z|)"]
) %>%
  filter(!grepl("Intercept", variable)) %>%
  mutate(
    # Extract outcome level from variable name
    outcome = case_when(
      grepl(":1$", variable) ~ "Prediabetes",
      grepl(":2$", variable) ~ "Diabetes",
      TRUE ~ "Unknown"
    ),
    # Clean variable name (remove :1 and :2 suffixes)
    var_original = gsub(":1$|:2$", "", variable),
    # Calculate confidence intervals
    ci_lower = estimate - 1.96 * std_error,
    ci_upper = estimate + 1.96 * std_error,
    # Significance levels
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      p_value < 0.1 ~ ".",
      TRUE ~ "NS"
    ),
    sig_factor = factor(significance, 
                       levels = c("***", "**", "*", ".", "NS"))
  )

# Separate main effects from interactions (if any)
coef_df <- coef_df %>%
  mutate(
    effect_type = ifelse(grepl(":", var_original), "Interaction", "Main Effect"),
    # Clean variable names for better display
    var_clean = gsub("Yes$", "", var_original),
    var_clean = gsub("High ", "High-", var_clean),
    var_clean = gsub("HeartDiseaseorAttack", "HeartDisease", var_clean)
  )

## PLOT 1: Coefficient Plot with Confidence Intervals (Main Effects)
main_effects <- coef_df %>% filter(effect_type == "Main Effect")

p1 <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value), 
                                y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                width = 0.3, position = position_dodge(width = 0.5)) +
  coord_flip() +
  facet_wrap(~outcome, ncol = 2) +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  labs(
    title = "Main Effects on Diabetes/Prediabetes Risk (Ordered by Significance)",
    subtitle = "Log-odds coefficients with 95% CI | Reference: No diabetes",
    x = "Predictor",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 11)
  )

print(p1)

## PLOT 2: Interaction Effects (if any exist)
interactions <- coef_df %>% filter(effect_type == "Interaction")

if (nrow(interactions) > 0) {
  p2 <- ggplot(interactions, aes(x = reorder(var_clean, -p_value), 
                                  y = estimate, color = sig_factor)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 3, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                  width = 0.3, position = position_dodge(width = 0.5)) +
    coord_flip() +
    facet_wrap(~outcome, ncol = 2) +
    scale_color_manual(
      values = c("***" = "#D55E00", "**" = "#E69F00", 
                 "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
      name = "Significance"
    ) +
    labs(
      title = "Interaction Effects on Diabetes/Prediabetes Risk",
      subtitle = "Log-odds coefficients with 95% CI",
      x = "Interaction Term",
      y = "Coefficient (log-odds)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 8),
      strip.text = element_text(face = "bold", size = 11)
    )
  
  print(p2)
} else {
  cat("No interaction terms in this model.\n")
  p2 <- NULL
}

## PLOT 3: Odds Ratios (easier interpretation)
main_effects <- main_effects %>%
  mutate(
    OR = exp(estimate),
    OR_lower = exp(ci_lower),
    OR_upper = exp(ci_upper)
  )

p3 <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value), 
                                y = OR, color = sig_factor)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = OR_lower, ymax = OR_upper), 
                width = 0.3, position = position_dodge(width = 0.5)) +
  coord_flip() +
  facet_wrap(~outcome, ncol = 2) +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  scale_y_continuous(trans = "log10", 
                     breaks = c(0.1, 0.5, 1, 2, 5, 10, 20)) +
  labs(
    title = "Odds Ratios for Main Effects (Ordered by Significance)",
    subtitle = "OR > 1: increased risk | OR < 1: decreased risk | Reference: No diabetes",
    x = "Predictor",
    y = "Odds Ratio (log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 11)
  )

print(p3)

## PLOT 4: Top 20 Most Significant Effects (combined across outcomes)
top20 <- coef_df %>%
  filter(effect_type == "Main Effect") %>%
  arrange(p_value) %>%
  group_by(var_clean) %>%
  slice_min(p_value, n = 1) %>%
  ungroup() %>%
  head(20) %>%
  mutate(
    OR = exp(estimate),
    direction = ifelse(estimate > 0, "Increases Risk", "Decreases Risk")
  )

p4 <- ggplot(top20, aes(x = reorder(var_clean, -p_value), 
                        y = estimate, fill = direction)) +
  geom_col() +
  geom_text(aes(label = significance), 
            hjust = ifelse(top20$estimate > 0, -0.3, 1.3),
            size = 3) +
  coord_flip() +
  facet_wrap(~outcome, ncol = 2) +
  scale_fill_manual(values = c("Increases Risk" = "#D55E00", 
                                "Decreases Risk" = "#0072B2")) +
  labs(
    title = "Top 20 Most Significant Predictors",
    subtitle = "Ranked by statistical significance",
    x = "Predictor",
    y = "Coefficient (log-odds)",
    fill = "Effect Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11)
  )

print(p4)

## PLOT 5: Odds Ratios for Interaction Effects (if applicable)
if (nrow(interactions) > 0) {
  interactions <- interactions %>%
    mutate(
      OR = exp(estimate),
      OR_lower = exp(ci_lower),
      OR_upper = exp(ci_upper)
    )
  
  p5 <- ggplot(interactions, aes(x = reorder(var_clean, -p_value), 
                                  y = OR, color = sig_factor)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    geom_point(size = 3, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = OR_lower, ymax = OR_upper), 
                  width = 0.3, position = position_dodge(width = 0.5)) +
    coord_flip() +
    facet_wrap(~outcome, ncol = 2) +
    scale_color_manual(
      values = c("***" = "#D55E00", "**" = "#E69F00", 
                 "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
      name = "Significance"
    ) +
    scale_y_continuous(trans = "log10") +
    labs(
      title = "Odds Ratios for Interaction Effects",
      subtitle = "OR > 1: synergistic effect | OR < 1: antagonistic effect",
      x = "Interaction Term",
      y = "Odds Ratio (log scale)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 8),
      strip.text = element_text(face = "bold", size = 11)
    )
  
  print(p5)
} else {
  p5 <- NULL
}

## Save all plots to files with appropriate sizes

# Plot 1: Main Effects - tall plot to accommodate many variables
ggsave("multinomial_plot1_main_effects.png", plot = p1, 
       width = 12, height = 16, dpi = 300, bg = "white")

# Plot 2: Interaction Effects (if exists)
if (!is.null(p2)) {
  ggsave("multinomial_plot2_interactions.png", plot = p2, 
         width = 12, height = 10, dpi = 300, bg = "white")
}

# Plot 3: Odds Ratios - tall plot for clarity
ggsave("multinomial_plot3_odds_ratios.png", plot = p3, 
       width = 12, height = 16, dpi = 300, bg = "white")

# Plot 4: Top 20 Most Significant
ggsave("multinomial_plot4_top20_significant.png", plot = p4, 
       width = 12, height = 10, dpi = 300, bg = "white")

# Plot 5: Interaction Odds Ratios (if exists)
if (!is.null(p5)) {
  ggsave("multinomial_plot5_interaction_odds_ratios.png", plot = p5, 
         width = 12, height = 10, dpi = 300, bg = "white")
}

# Optional: Save as PDF for publication quality
ggsave("multinomial_plot1_main_effects.pdf", plot = p1, 
       width = 12, height = 16)

if (!is.null(p2)) {
  ggsave("multinomial_plot2_interactions.pdf", plot = p2, 
         width = 12, height = 10)
}

ggsave("multinomial_plot3_odds_ratios.pdf", plot = p3, 
       width = 12, height = 16)

ggsave("multinomial_plot4_top20_significant.pdf", plot = p4, 
       width = 12, height = 10)

if (!is.null(p5)) {
  ggsave("multinomial_plot5_interaction_odds_ratios.pdf", plot = p5, 
         width = 12, height = 10)
}

cat("\n✓ All plots generated and saved successfully!\n")
```


```{r}
main_effects <- coef_df %>%
  filter(effect_type == "Main Effect", p_value < 0.05)  #Only significant

p1_nn <- ggplot(main_effects, aes(x = reorder(var_clean, -p_value), 
                                y = estimate, color = sig_factor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                width = 0.3, position = position_dodge(width = 0.5)) +
  coord_flip() +
  facet_wrap(~outcome, ncol = 2) +
  scale_color_manual(
    values = c("***" = "#D55E00", "**" = "#E69F00", 
               "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
    name = "Significance"
  ) +
  labs(
    title = "Main Effects on Diabetes/Prediabetes Risk (Ordered by Significance)",
    subtitle = "Log-odds coefficients with 95% CI | Reference: No diabetes",
    x = "Predictor",
    y = "Coefficient (log-odds)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 11)
  )

print(p1_nn)

ggsave("m1.png", plot = p1_nn, 
       width = 12, height = 10, dpi = 300, bg = "white")

```

```{r}

## PLOT 2: Interaction Effects (if any exist)
interactions <- coef_df %>%
  filter(effect_type == "Interaction", p_value < 0.05)  # only significant


p2_nn <- ggplot(interactions, aes(x = reorder(var_clean, -p_value), 
                                  y = estimate, color = sig_factor)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 3, position = position_dodge(width = 0.5)) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                  width = 0.3, position = position_dodge(width = 0.5)) +
    coord_flip() +
    facet_wrap(~outcome, ncol = 2) +
    scale_color_manual(
      values = c("***" = "#D55E00", "**" = "#E69F00", 
                 "*" = "#56B4E9", "." = "#009E73", "NS" = "#CCCCCC"),
      name = "Significance"
    ) +
    labs(
      title = "Interaction Effects on Diabetes/Prediabetes Risk",
      subtitle = "Log-odds coefficients with 95% CI",
      x = "Interaction Term",
      y = "Coefficient (log-odds)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 8),
      strip.text = element_text(face = "bold", size = 11)
    )
  
  print(p2_nn)
  
  ggsave("m2.png", plot = p2_nn, 
         width = 12, height = 10, dpi = 300, bg = "white")
```
```{r}

top20 <- coef_df %>%
  arrange(p_value) %>%
  group_by(outcome, var_clean) %>%   # group by both
  slice_min(p_value, n = 1) %>%
  ungroup() %>%
  filter(p_value < 0.05) %>%
  group_by(outcome) %>%
  slice_head(n = 20) %>%             # top 20 per outcome
  ungroup() %>%
  mutate(
    OR = exp(estimate),
    direction = ifelse(estimate > 0, "Increases Risk", "Decreases Risk")
  )

# Helper function for reordering within facets
reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
  new_x <- paste(x, within, sep = sep)
  stats::reorder(new_x, by, FUN = fun)
}

scale_x_reordered <- function(..., sep = "___") {
  reg <- paste0(sep, ".+$")
  ggplot2::scale_x_discrete(labels = function(x) gsub(reg, "", x), ...)
}

p4 <- ggplot(top20, aes(x = reorder_within(var_clean, -p_value, outcome), 
                        y = estimate, fill = direction)) +
  geom_col() +
  geom_text(aes(label = significance,
                hjust = ifelse(estimate > 0, -0.3, 1.3)),
            size = 2.5) +
  coord_flip() +
  facet_wrap(~outcome, ncol = 2, scales = "free_y") +
  scale_x_reordered() +
  scale_fill_manual(values = c("Increases Risk" = "#D55E00", 
                               "Decreases Risk" = "#0072B2")) +
  labs(
    title = "Top 20 Most Significant Predictors by Outcome",
    subtitle = "Ranked by statistical significance within each outcome",
    x = "Predictor",
    y = "Coefficient (log-odds)",
    fill = "Effect Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.y = element_text(size = 8)
  )

print(p4)

ggsave("T2.png", plot = p4, 
         width = 12, height = 10, dpi = 300, bg = "white")

```


```{r}
library(pROC)
rocplot <- roc(Diabetes_binary ~ fitted(fit1), data=dbts_df)
plot.roc(rocplot, legacy.axes=TRUE) # Specficity on x axis if legacy.axes=F
auc(rocplot) # auc = area under ROC curve = concordance index
```


```{r}
library(MASS)
stepAIC(fit1)
```

```{r}
library(car)
Anova(fit4)
influencePlot(fit1,id = T)
```
```{r}
```


```{r}
dbts_df[c(76213,127185,147377,203560,209993,212800),]
```
```{r}
pred.prob <- fitted(fit1)
pred.prob 
```
```{r}
lp <- predict(fit1, se.fit=TRUE) # linear predictor
LB <- lp$fit - 1.96*lp$se.fit # confidence bounds for linear predictor
UB <- lp$fit + 1.96*lp$se.fit # better: use qnorm(0.975) instead of 1.96
LB.p <- exp(LB)/(1 + exp(LB)) # confidence bounds for P(Y=1)
UB.p <- exp(UB)/(1 + exp(UB))
cbind(pred.prob, LB.p, UB.p)
```


```{r}
library(VGAM)

fit5 <- vglm(Diabetes_012~HighBP + HighChol + CholCheck + BMI + Smoker + 
    Stroke + HeartDiseaseorAttack + PhysActivity + Fruits + Veggies + 
    HvyAlcoholConsump + AnyHealthcare + NoDocbcCost + GenHlth + 
    MentHlth + PhysHlth + DiffWalk + Sex + Age + Education + 
    Income, data = dbts_df, family = multinomial)
summary(fit5)
```


